Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-07 talk/icooolps2009/paper.tex
small fixes, and rephrasing some sentences that I found to heavyweight

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot loop
of the language interpreter is its bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to a \emph{user loop}. User loops
occur when the program counter of the \emph{language interpreter} has the
same value several times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pair} variables.

@@

In addition to the classification of the variables, there are two methods of
\texttt{JitDriver} that need to be called. Both of them receive as arguments the
current values of the variables listed in the definition of the driver. The
first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
of the body of the bytecode dispatch loop. The other, more interesting one, is
\texttt{can\_enter\_jit}. This method needs to be called at the end of any
instruction that can set the program counter of the language interpreter to an
earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
instruction, and only if it is actually a backward jump. The place where this
method is called is where the language interpreter performs profiling to decide
when to start tracing. It is also the place where the tracing JIT checks
whether a loop is closed. This is considered to be the case when the values of
the ``green'' variables are the same as at an earlier call to the
\texttt{can\_enter\_jit} method.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It should be possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
that when no hints were present in the interpreter at all.

@@

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 46340 (the smallest number whose square still fits into a 32
bit word) using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

@@

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked to measure the overhead of the profiling. For this interpreter
it to be rather large, with 50\% slowdown due to profiling. This is
because the example interpreter needs to do one hash table lookup per loop
iteration. For larger interpreters (e.g. the Python one) it seems likely that
the overhead is less significant, given that many operations in Python need
hash-table lookups themselves.

>>>>

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\vspace{-0.4cm}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

To improve this situation, the tracing JIT could trace the execution of several
opcodes, thus effectively unrolling the bytecode dispatch loop. Ideally, the
bytecode dispatch loop should be unrolled exactly so much, that the unrolled version
corresponds to a \emph{user loop}. User loops
occur when the program counter of the \emph{language interpreter} has the
same value several times. This program counter is typically stored in one or several
variables in the language interpreter, for example the bytecode object of the
currently executed function of the user program and the position of the current
bytecode within that.  In the example above, the program counter is represented by 
the \texttt{bytecode} and \texttt{pc} variables.

@@

In addition to the classification of the variables, there are two methods of
\texttt{JitDriver} that need to be called. Both of them receive as arguments the
current values of the variables listed in the definition of the driver. The
first one is \texttt{jit\_merge\_point} which needs to be put at the beginning
of the body of the bytecode dispatch loop. The other, more interesting one, is
\texttt{can\_enter\_jit}. This method needs to be called at the end of any
instruction that can set the program counter of the language interpreter to an
earlier value. For the example this is only the \texttt{JUMP\_IF\_A}
instruction, and only if it is actually a backward jump. Here is
where the language interpreter performs profiling to decide
when to start tracing. It is also the place where the tracing JIT checks
whether a loop is closed. This is considered to be the case when the values of
the ``green'' variables are the same as at an earlier call to the
\texttt{can\_enter\_jit} method.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It is possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
that when no hints were present in the interpreter at all.

@@

The first round of benchmarks (Figure \ref{fig:bench1}) are timings of the
example interpreter (Figure \ref{fig:tlr-basic}) used in this paper computing
the square of 46340 (the largest number whose square still fits into a 32
bit word) using the bytecode of Figure \ref{fig:square}. The results for various
constellations are as follows:

@@

\textbf{Benchmark 5:} Same as before, but with the threshold set so high that the tracer is
never invoked to measure the overhead of the profiling. For this interpreter
it to be rather large, with 50\% slowdown due to profiling. This is because the interpreter 
is small and the opcodes simple. For larger interpreters (e.g. the Python one) it seems 
likely that the overhead is less significant.
